A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-14 paper.tex
Updates from Overleaf

The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, some definitions of ARGs (especially those linked to a specific stochastic process) are overly narrow, which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we 
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] that persist over 
many generations. A ``genome'' in our usage corresponds to the 
haploid genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the graph connectivity 
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome 
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges. Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.

>>>>

The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly some definitions of ARGs, especially those linked to a specific stochastic process, are overly narrow which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations. A ``genome'' in our usage corresponds to the
haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

The topology of a gARG specifies that genetic inheritance occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges. Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.
